---
title: "Johns Hopkins Data Science Specialization : Milestone Report"
subtitle : "Exploratory Data Analysis and Modeling"
author : "Suhas. P. K"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "dark"
    downcute_theme: "chaos"
---


```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```

## Project Overview

### Introduction

Introduction to mobile devices led many inventions and innovations. Once there was a time where sending an email was a luxury. Within a span of two decades the evolution of communication devices from a huge hefty computing machines to a simple palm held, touch responsive mobile phone is observed significantly.

One of the basic features of these mobile devices is to predicts words based on the sentences that the user types.

In this **milestone project**, with the help of **SwiftKey** and **Johns Hopkins Data Science Specialization Course**, I will be attempting to build a predictive text models like those used by SwiftKey.

In this project includes analyzing a large corpus of text documents to discover the structure in the data, and how words are put
together. This involves cleaning and analysisng text data, building and sampling predictive text model. And build a predictive text product using Rshiny.

### Resources 

- [Text mining infrastructure in R](https://www.jstatsoft.org/article/view/v025i05)
- [CRAN Task View: Natural Language Processing](https://cran.r-project.org/web/views/NaturalLanguageProcessing.html)
- [NLP slides](https://web.stanford.edu/~jurafsky/NLPCourseraSlides.html)
-[NLP Wiki](https://en.wikipedia.org/wiki/Natural_language_processing)

### Course datasets

This is the training dataset that I will be using for the basic exercises of this project.

[Download here](https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip)

## Game Plan

### Libraries

```{r libraries, message=FALSE}
if(!require(knitr)){
  install.packages("knitr")
  library(knitr)
}
if(!require(stringi)){
  install.packages("stringi")
  library(stringi)
}
if(!require(NLP)){
  install.packages("NLP")
  library(NLP)
}
if(!require(tm)){
  install.packages("tm")
  library(tm)
}
if(!require(RWeka)){
  install.packages("RWeka")
  library(RWeka)
}
if(!require(data.table)){
  install.packages("data.table")
  library(data.table)
}
if(!require(ggplot2)){
  install.packages("ggplot2")
  library(ggplot2)
}
if(!require(SnowballC)){
  install.packages("SnowballC")
  library(SnowballC)
}
if(!require(wordcloud)){
  install.packages("wordcloud")
  library(wordcloud)
}
if(!require(kableExtra)){
  install.packages("kableExtra")
  library(kableExtra)
}
if(!require(gridExtra)){
  install.packages("gridExtra")
  library(gridExtra)
}
```

### Importing the datasets.
```{r import datasets,message=FALSE}

# Read Blog lines
blogs <- file(
  paste(getwd(),'Coursera-SwiftKey/final/en_US/en_US.blogs.txt',sep = "/"),
  "r"
)
blog_lines <- readLines(blogs, warn = FALSE, encoding = "UTF-8", skipNul = TRUE)
close(blogs)

# Read News lines
news <- file(
  paste(getwd(),'Coursera-SwiftKey/final/en_US/en_US.news.txt',sep = "/"),
  "r"
)
news_lines <- readLines(news,warn = FALSE, encoding = "UTF-8", skipNul = TRUE)
close(news)

# Read Twitter Lines
twitter <- file(
  paste(getwd(),'Coursera-SwiftKey/final/en_US/en_US.twitter.txt',sep = "/"),
  "r"
)
twitter_lines <- readLines(twitter,warn = FALSE, encoding = "UTF-8", skipNul = TRUE)
close(twitter)


```

### General statistics on 'Words per line (wpl)'

```{r wpl, message=FALSE}

# number of lines 
num_lines <- sapply(list(blog_lines,news_lines,twitter_lines), length)

# number of characters
num_char <- sapply(list(nchar(blog_lines),nchar(news_lines),nchar(twitter_lines)),sum)

# number of words 
num_words <- sapply(list(blog_lines,news_lines,twitter_lines), stri_stats_latex)[4,]

# number of words per line
wpl <- lapply(list(blog_lines,news_lines,twitter_lines), function(x){stri_count_words(x)})

# summary 
wpl_gist <- sapply(list(blog_lines,news_lines,twitter_lines), function(x){
  summary(stri_count_words(x))[c('Min.', 'Mean', 'Max.')]})

rownames(wpl_gist) = c('wpl_min', 'wpl_mean', 'wpl_max')

gist <- data.frame(
  FileName = c("en_US.blogs.txt","en_US.news.txt","en_US.twitter.txt"),
  FileSize = sapply(
    list(blog_lines,news_lines,twitter_lines), function(x){format(object.size(x),"MB")}
  ),
  Lines = num_lines,
  Charaters = num_char,
  Words = num_words,
  t(rbind(round(wpl_gist)))
  
)

# Tabular representation

gist %>% kbl() %>%  kable_styling(bootstrap_options = c("condensed", "responsive"))

```



```{r wpl_histogram, message=FALSE, warning=FALSE}
plot1 <- qplot(wpl[[1]],
               geom = "histogram",
               main = "US Blogs",
               xlim = c(0,200),
               xlab = "Words per Line",
               ylab = "Frequency",
               binwidth = 5)

plot2 <- qplot(wpl[[2]],
               geom = "histogram",
               main = "US News",
               xlim = c(0,300),
               xlab = "Words per Line",
               ylab = "Frequency",
               binwidth = 5)

plot3 <- qplot(wpl[[3]],
               geom = "histogram",
               main = "US Twitter",
               xlab = "Words per Line",
               ylab = "Frequency",
               binwidth = 1)

plotList = list(plot1, plot2, plot3)
do.call(grid.arrange, c(plotList, list(ncol = 1)))

# free up some memory
rm(plot1, plot2, plot3)
```

```{r sampling and cleaning, message=FALSE}
set.seed(20042)

# assigning sample size
sampleSize = 0.01
# sampling all 3 datasets
sample_blogs <- sample(blog_lines, length(blog_lines)*sampleSize, replace = FALSE)

sample_news <- sample(news_lines, length(blog_lines)*sampleSize, replace = FALSE)

sample_twitter <- sample(twitter_lines, length(blog_lines)*sampleSize, replace = FALSE)

# Removing all non-English characters from the sample data.

sample_blogs <- iconv(sample_blogs, "latin1", "ASCII", sub="")

sample_news <- iconv(sample_news, "latin1", "ASCII", sub="")

sample_twitter <- iconv(sample_twitter, "latin1", "ASCII", sub="")

# Combining all 3 sample data sets into a single data set and writing to disk.

sample_data <- c(sample_blogs,sample_news,sample_twitter)
sample_data_filename <- paste(getwd(),                          "Coursera-SwiftKey/final/en_US/en_US.sample.txt",sep = "/")
con <- file(sample_data_filename, open = "w")
writeLines(sample_data,con)
close(con)

# number of lines and words in the sample data set.
sample_data_lines <- length(sample_data)
sample_data_words <- sum(stri_count_words(sample_data))

```









